### 2022-06-21 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@eea6e201f3438c201940b84fa06464c6f96cb8a2 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover risk,\textsuperscript{2,3} these
approaches oversimplify the relevant interspecific heterogeneity in
immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{4} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{5} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts {[}Geoghegan{]}. This creates a disconnect
between coevolutionary theory (including empirical evidence from
virology) and most existing ecological frameworks for mapping spillover
risk.

The Geographic Mosaic Theory of Coevolution\textsuperscript{GMTC; 6}
attempts to explicitly connect microevolutionary dynamics to the
macroecology and biogeography of symbiotic interactions. The GMTC posits
that coevolutionary processes among pairs\textsuperscript{7} or
complexes\textsuperscript{8} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanism,
giving rise to fragmented systems with different structure and
ecologically dynamics over large spatial extents.\textsuperscript{see
\emph{e.g.} 9} The GMTC predicts a spatial fragmentation of
coevolutionary dynamics under the joint action of three
processes:\textsuperscript{see notably 10} coevolutionary hot- and
coldspots, which appear when the intensity of \emph{interaction} (in
terms of reciprocal fitness consequences) varies spatially; selection
mosaics, wherein the intensity of \emph{selection} varies across space,
driven by both the biotic complexity of the community (locally diverse
hosts and viruses are more biotically complex) and the local
favorability of the environment;\textsuperscript{11} and trait remixing,
which occurs when coevolutionary dynamics are driven by by the arrival
(or departure) of \emph{functional traits}, through changes in community
composition due to invasions, meta-community dynamics, and disperal.

@@

Bats, the second most diverse groups of mammals, are found worldwide and
serve as the main animal reservoir for different strains of
betacoronaviruses.\textsuperscript{12} This has attracted attention to
areas where high diversity of bats, and therefore presumably high
diversity of betacoronaviruses, can be an important issue for human
health.\textsuperscript{13,14} By overlaying the IUCN range maps for
confirmed bat hosts of betacoronaviruses {[}fig.~\ref{fig:richness};
top{]}, we see that the the main hotspots (both in terms of size and
higher values) of host richness are primarily South-Eastern Asia, parts
of Southern Europe, and to a lesser extent parts of Africa in the -25-0
range of latitudes. The description of host richness is an important
first step towards understanding risk, as previous
research\textsuperscript{2,15} states that locally diverse bat
communities could maintain more viruses and hence, a higher probability
of having a pathogen that could represent a risk for human health.

@@

Nevertheless, locally diverse and virus-rich bat communities could
represent an increased risk of spillover under climate change through
the creation of novel interactions,\textsuperscript{16} and therefore
the diversity of \emph{Betacoronavirus} strains should similarly be
ccounted for. In fig.~\ref{fig:richness} (bottom), we contrast the
evolutionary distinctiveness of bats and viruses -- this reveals a
slightly different portrait than bat richness alone. Chiropterans can be
classified, from a macro-evolutionary standpoint, as Yangochiroptera and
Yinpterochiroptera elsewhere.\textsuperscript{17,18} Specifically, we
would expect that the so-called ``New World'' group of bats, being more
evolutionary distinct, would also have evolutionary distinct viruses.
Indeed fig.~\ref{fig:richness} (bottom) reveals it to be the case, and
this region harbors a distinct bat-betacoronaviruses complex. This can
be explained by the fact that Yangochiroptera, although not limited to
the western hemisphere, contain the highly diverse adaptive radiation in
the Phyllostomidae,\textsuperscript{19} which is restricted to the
western hemisphere. By contrast, South-Eastern Asia has a lot of
non-evolutionary distinct bats, who nevertheless hosted
evolutionary-distinct viruses.

It is noteworthy that outside of South America, viral evolutionary
distinctiveness does not accurately track host diversity, with some
areas having over-distinct viruses (eastern China but, oddly, not the
rest of southeast Asia). There are a number of likely explanations.
First, given the richness of bats in southeast Asia, many
betacoronaviruses likely remain to be discovered in this region. Indeed,
global predictions by\textsuperscript{20} highlight that southeast Asia
is a likely hostpot of unconfirmed hosts of betacoronaviruses, which
would likely result in additional viral discoveries. This idea is
unsurprising given the growing realization, especially since the
emergence of SARS-CoV-2, that unique lineages of similar viruses are
widespread in bats but still mostly undescribed. The most distinct
bats-betacoronaviruses complex is found in South America, a region with
a comparatively lower number of hosts; this matches with the isolation
through viariance of the host group, and may highlight a different
co-evolutionary dynamic. Alternatively, this distinctiveness hostpot may
be a product of under-sampling: South-America is one of the places where
the fewest \emph{Betacoronavirus} sequences have been
discovered,\textsuperscript{2,21,22} resulting in sparser phylogenetic
tree, thereby artificially inflating distinctiveness. Adding more
viruses would bring the distinctiveness of known sequences down.
Finally, South-America is the range of Phyllostomidae, a group of bats
that underwent explosive diversification events,\textsuperscript{19}
which may drive the emergence of multiple viral lineages.

@@

Despite the difference in evolutionary distinctiveness globally, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{23,24} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{25} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{26,27} in addition to being hardcoded in their
genome,\textsuperscript{28} may be related to a wide variety of
diets,\textsuperscript{25,29} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{30}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{31} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{32} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{33} and Europe\textsuperscript{34} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

As host richness, joint distinctiveness, or phylogeographic structure
suggest that the bat-betacoronaviruses complex is globally fragmented
enough to give rise to both different levels of risk (as evidenced by
the spatial location of spillover events) and different types of
co-evolutionary dynamics, we turn to the Geographic Mosaic Theory of
Coevolution\textsuperscript{6} to provide a measure of risk accounting
for multiple processes. In fig.~\ref{fig:trivariate}, we overlapped
three components of spillover risk: viral sharing, \emph{i.e.} the
chance that two bats will share viruses overall; Local Contribution to
Beta Diversity, \emph{i.e.} the fact that a bat community is
compositionally unique compared to the average compositional similarity
across the entire system; finally, host phylogenetic diversity,
\emph{i.e.} how dispersed the bats in a location are within the tree of
life. This approach leads to the definition of broad biogeographic
regions of risk, where the same color represents the same type of risk.
By way of constrat to figures fig.~\ref{fig:richness} and
fig.~\ref{fig:biogeo}, these regions do not necessarilly overlap with
previous spatial partitions of the bat-betacoronaviruses complex.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{\emph{e.g.} 35} Recent
surveillance\textsuperscript{36} has identified a novel
\emph{Betacoronavirus} (in the subgenus \emph{Nobecovirus}) in
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus},
\emph{Rousettus madagascariensis}), emphasizing strong proof of
principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{37} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{38} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of risk overlays
with recent results from\textsuperscript{39} -- areas of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus}, stressing the need for
spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{23,24} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{40,41}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{42} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{43} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{44,45} and
vectors of pathogens that put a risk on human
health.\textsuperscript{46,47} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{22} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{20} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{48} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{49} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{20,39} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{16}

{[}Tim tinker with para about conservation{]} There are several factors
that drive changes in the diversity of bats,\textsuperscript{50} but
human activities' effects on the ecosystem (like modifications of land
use) could significantly decrease it. Therefore, it can be suggested
that changes in the diversity of betacoronaviruses in bats are linked to
their biogeographic variation, and human population density and other
anthropogenic factors are decisive moderators for its implications in
public health. With the increase of contact between humans and potential
hosts, we also increase the risk of emergence of novel
diseases,\textsuperscript{51} as previous studies on RNA viruses suggest
the importance of host phylogeography at the time of virus
dispersal.\textsuperscript{52} One of these scenarios where interaction
between bats and humans can occur can be seed dispersal in tropical
agroecosystems. It opens the discussion of whether the fruits thrown by
bats not only disperse seeds but could also be a source of indirect
interaction between viruses of bat origin and
humans.\textsuperscript{53} This represents a challenge for conservation
strategies and disease ecology since some areas can haveboth potential
for the acquisition of zoonotic viruses and bat-human interactions; in
particular, the challenge lies in the fact that actual exposure must
then be quantified accounting for several transmission scenarios,
including both direct and indirect bat - human interaction.

@@

We downloaded the data on bats hosts of \emph{Betacoronavirus} assembled
by\textsuperscript{20} from
\texttt{https://www.viralemergence.org/betacov} on Apr.~2022, and
filtered it to ``known'' hosts (established before the emergence of
SARS-CoV-2) and ``novel'' hosts (confirmed through sampling and
competence assays since the initial data collection). The original
database was assembled by a combination of data mining and literature
surveys, including automated alerts on the ``bats'' and ``coronavirus''
keywords to identify novel empirical evidence of bats-betacoronaviruses
associations; this yielded a total of 126 known hosts, 47 of which were
novel hosts.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{54} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{55} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{56} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{57} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range
overlap.\textsuperscript{61} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover risk,\textsuperscript{2,3} these
approaches oversimplify the relevant interspecific heterogeneity in
immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{4} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{5} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{6} This creates a disconnect
between coevolutionary theory (including empirical evidence from
virology) and most existing ecological frameworks for mapping spillover
risk.

The Geographic Mosaic Theory of Coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{7} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{8} or
complexes\textsuperscript{9} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanism,
giving rise to fragmented systems with different structure and
ecologically dynamics over large spatial extents.\textsuperscript{10}
The GMTC predicts a spatial fragmentation of coevolutionary dynamics
under the joint action of three processes:\textsuperscript{11}
coevolutionary hot- and coldspots, which appear when the intensity of
\emph{interaction} (in terms of reciprocal fitness consequences) varies
spatially; selection mosaics, wherein the intensity of \emph{selection}
varies across space, driven by both the biotic complexity of the
community (locally diverse hosts and viruses are more biotically
complex) and the local favorability of the
environment;\textsuperscript{12} and trait remixing, which occurs when
coevolutionary dynamics are driven by by the arrival (or departure) of
\emph{functional traits}, through changes in community composition due
to invasions, meta-community dynamics, and disperal.

@@

Bats, the second most diverse groups of mammals, are found worldwide and
serve as the main animal reservoir for different strains of
betacoronaviruses.\textsuperscript{13} This has attracted attention to
areas where high diversity of bats, and therefore presumably high
diversity of betacoronaviruses, can be an important issue for human
health.\textsuperscript{14,15} By overlaying the IUCN range maps for
confirmed bat hosts of betacoronaviruses {[}fig.~\ref{fig:richness};
top{]}, we see that the the main hotspots (both in terms of size and
higher values) of host richness are primarily South-Eastern Asia, parts
of Southern Europe, and to a lesser extent parts of Africa in the -25-0
range of latitudes. The description of host richness is an important
first step towards understanding risk, as previous
research\textsuperscript{2,16} states that locally diverse bat
communities could maintain more viruses and hence, a higher probability
of having a pathogen that could represent a risk for human health.

@@

Nevertheless, locally diverse and virus-rich bat communities could
represent an increased risk of spillover under climate change through
the creation of novel interactions,\textsuperscript{17} and therefore
the diversity of \emph{Betacoronavirus} strains should similarly be
ccounted for. In fig.~\ref{fig:richness} (bottom), we contrast the
evolutionary distinctiveness of bats and viruses -- this reveals a
slightly different portrait than bat richness alone. Chiropterans can be
classified, from a macro-evolutionary standpoint, as Yangochiroptera and
Yinpterochiroptera elsewhere.\textsuperscript{18,19} Specifically, we
would expect that the so-called ``New World'' group of bats, being more
evolutionary distinct, would also have evolutionary distinct viruses.
Indeed fig.~\ref{fig:richness} (bottom) reveals it to be the case, and
this region harbors a distinct bat-betacoronaviruses complex. This can
be explained by the fact that Yangochiroptera, although not limited to
the western hemisphere, contain the highly diverse adaptive radiation in
the Phyllostomidae,\textsuperscript{20} which is restricted to the
western hemisphere. By contrast, South-Eastern Asia has a lot of
non-evolutionary distinct bats, who nevertheless hosted
evolutionary-distinct viruses.

It is noteworthy that outside of South America, viral evolutionary
distinctiveness does not accurately track host diversity, with some
areas having over-distinct viruses (eastern China but, oddly, not the
rest of southeast Asia). There are a number of likely explanations.
First, given the richness of bats in southeast Asia, many
betacoronaviruses likely remain to be discovered in this region. Indeed,
global predictions highlight that southeast Asia is a likely hostpot of
unconfirmed hosts of betacoronaviruses,\textsuperscript{21} which would
likely result in additional viral discoveries. This idea is unsurprising
given the growing realization, especially since the emergence of
SARS-CoV-2, that unique lineages of similar viruses are widespread in
bats but still mostly undescribed. The most distinct
bats-betacoronaviruses complex is found in South America, a region with
a comparatively lower number of hosts; this matches with the isolation
through viariance of the host group, and may highlight a different
co-evolutionary dynamic. Alternatively, this distinctiveness hostpot may
be a product of under-sampling: South-America is one of the places where
the fewest \emph{Betacoronavirus} sequences have been
discovered,\textsuperscript{2,22,23} resulting in sparser phylogenetic
tree, thereby artificially inflating distinctiveness. Adding more
viruses would bring the distinctiveness of known sequences down.
Finally, South-America is the range of Phyllostomidae, a group of bats
that underwent explosive diversification events,\textsuperscript{20}
which may drive the emergence of multiple viral lineages.

@@

Despite the difference in evolutionary distinctiveness globally, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{24,25} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{26} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{27,28} in addition to being hardcoded in their
genome,\textsuperscript{29} may be related to a wide variety of
diets,\textsuperscript{26,30} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{31}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{32} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{33} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{34} and Europe\textsuperscript{35} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

As host richness, joint distinctiveness, or phylogeographic structure
suggest that the bat-betacoronaviruses complex is globally fragmented
enough to give rise to both different levels of risk (as evidenced by
the spatial location of spillover events) and different types of
co-evolutionary dynamics, we turn to the Geographic Mosaic Theory of
Coevolution\textsuperscript{7} to provide a measure of risk accounting
for multiple processes. In fig.~\ref{fig:trivariate}, we overlapped
three components of spillover risk: viral sharing, \emph{i.e.} the
chance that two bats will share viruses overall; Local Contribution to
Beta Diversity, \emph{i.e.} the fact that a bat community is
compositionally unique compared to the average compositional similarity
across the entire system; finally, host phylogenetic diversity,
\emph{i.e.} how dispersed the bats in a location are within the tree of
life. This approach leads to the definition of broad biogeographic
regions of risk, where the same color represents the same type of risk.
By way of constrat to figures fig.~\ref{fig:richness} and
fig.~\ref{fig:biogeo}, these regions do not necessarilly overlap with
previous spatial partitions of the bat-betacoronaviruses complex.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{\emph{e.g.} 36} Recent
surveillance\textsuperscript{37} has identified a novel
\emph{Betacoronavirus} (in the subgenus \emph{Nobecovirus}) in
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus},
\emph{Rousettus madagascariensis}), emphasizing strong proof of
principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{38} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{39} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of risk overlays
with recent results from\textsuperscript{40} -- areas of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus}, stressing the need for
spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{24,25} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{41,42}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{43} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{44} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{45,46} and
vectors of pathogens that put a risk on human
health.\textsuperscript{47,48} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{23} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{21} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{49} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{50} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{21,40} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{17}

{[}Tim tinker with para about conservation{]} There are several factors
that drive changes in the diversity of bats,\textsuperscript{51} but
human activities' effects on the ecosystem (like modifications of land
use) could significantly decrease it. Therefore, it can be suggested
that changes in the diversity of betacoronaviruses in bats are linked to
their biogeographic variation, and human population density and other
anthropogenic factors are decisive moderators for its implications in
public health. With the increase of contact between humans and potential
hosts, we also increase the risk of emergence of novel
diseases,\textsuperscript{52} as previous studies on RNA viruses suggest
the importance of host phylogeography at the time of virus
dispersal.\textsuperscript{53} One of these scenarios where interaction
between bats and humans can occur can be seed dispersal in tropical
agroecosystems. It opens the discussion of whether the fruits thrown by
bats not only disperse seeds but could also be a source of indirect
interaction between viruses of bat origin and
humans.\textsuperscript{54} This represents a challenge for conservation
strategies and disease ecology since some areas can haveboth potential
for the acquisition of zoonotic viruses and bat-human interactions; in
particular, the challenge lies in the fact that actual exposure must
then be quantified accounting for several transmission scenarios,
including both direct and indirect bat - human interaction.

@@

We downloaded the data on bats hosts of \emph{Betacoronavirus} assembled
by\textsuperscript{21} from
\texttt{https://www.viralemergence.org/betacov} on Apr.~2022, and
filtered it to ``known'' hosts (established before the emergence of
SARS-CoV-2) and ``novel'' hosts (confirmed through sampling and
competence assays since the initial data collection). The original
database was assembled by a combination of data mining and literature
surveys, including automated alerts on the ``bats'' and ``coronavirus''
keywords to identify novel empirical evidence of bats-betacoronaviruses
associations; this yielded a total of 126 known hosts, 47 of which were
novel hosts.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{55} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{56} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{57} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{58} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range
overlap.\textsuperscript{62} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.
